Trion ARPES signature

### 2024-02-14 paper/main.tex
update eeh trion

In most driven ARPES settings, the probe is much weaker than the pump 
and does not itself influence the electronic structure,
and the generic theory of ARPES can be formulated by 
treating the pump as a non-equilibrium self-energy, 
while the effect of the probe can be captured by the first order perturbation  
and hence the final ARPES intensity is proportional to the electron dipole and 
the pumped single-electron lesser Green function
\cite{freericks2009theoretical,schuler2021theory};
the theory reduced to the usual Fermi golden rule 
when the pump is turned off \cite{sobota2021angle,freericks2021two}.
When the pump pulse is strong at its peak but has faded away at the probing time, 
the main effect of pumping is preparing an initial state, 
and the ARPES intensity can be evaluated from 
pure-state first-order time dependent perturbation theory,
which gives a Fermi golden rule-like expression of the probability of an electron being pumped out of the sample~\cite{rustagi2018photoemission}.
Ref.~\cite{rustagi2018photoemission} is focused on excitons;
here we generalize its methodology to all excitations and have  
\begin{widetext}
    \begin{equation}
        I_{\vb{k}}(\omega, t) \propto \sum_{n_1, n_2} \sum_{n}
        \rho_n (M^{fn_2}_{\vb{k} \vb{k}_\parallel})^* M^{fn_1}_{\vb{k} \vb{k}_\parallel} 
        \int_{t_0}^t \dd{t_1} \int_{t_0}^t \dd{t_2}
        \ee^{-\ii (E_n - \omega ) (t_1 - t_2)}
        \mel{\Psi_n}{c^\dagger_{n_2 \vb{k}_\parallel} 
        U(t_2, t_1) c_{n_1 \vb{k}_\parallel}}{\Psi_n} s(t_1) s(t_2),
        \label{eq:exciton-response}
    \end{equation}
\end{widetext}
where $t$ is the observation time, 
$t_0$ is the starting time of the probe pulse,
$s(t)$ is the envelope function of the probe,
$\omega$ is the output electron energy shifted by the frequency of the probe pulse 
and the work function of the sample, 
$\vb{k}$ is the momentum of the outgoing electron 
and $\vb{k}_\parallel$ is its projection on the two-dimensional plane,  
$n_{1, 2}$ are electron band indices, 
$n = (S, \vb{Q})$ is the index of exciton modes, 
$\rho_n$ is the effective density matrix of excitons, 
$\ket*{\Psi_n}$ is the many-body wave function of the exciton state at $t=t_0$,
and $U(t_2, t_1)$ is the time evolution operator.
The $c^\dagger U c \ee^{- \ii E_n t}$ 
factor comes from the electron lesser Green function; 
the $\ee^{- \ii E_n t}$ factor comes from the time evolution of the excitation states, 
while the time evolution operator $U(t_2, t_1)$ 
instructs the time evolution of the residue of the excitation after one electron is driven out; 
time oscillation from the two factors together gives us the energy conservation relation
\begin{equation}
    \omega = E_n - E_{\text{remain}} , 
\end{equation}
where $E_{\text{residue}}$ refers to the remaining energy of the system after
one electron is driven out from $n_1, \kp$.
For three-dimensional systems, the momenta of $c$ and $c^\dagger$ in the $c^\dagger U c$ factor can be different \cite{freericks2009theoretical},  
but for two-dimensional materials however, since by momentum conservation  
the crystal momentum of the band electron that is driven out by the external field 
can only be the parallel component of $\vb{k}$, and we refer it as $\vb{k}_\parallel$ in \eqref{eq:exciton-response}.
As below we are mostly dealing with the ARPES spectrum of a single mode
In principle, the $\mel{\Psi_n}{c^\dagger_{n_2 \vb{k}_\parallel} U(t_2, t_1) c_{n_1 \vb{k}_\parallel}}{\Psi_n}$ matrix can be non-diagonal with an excited initial state;
frequently, however, because of energy conservation, the signature at given $(\omega, \vb{k})$
can only come from a single band, and in this case $n_1 = n_2$.

@@

We can define the canonical momenta of $\vb{r}_{1,2}$, 
hereafter referred to as $\vb{k}_{1,2}$, 
as $- \ii \hbar_{\vb{r}_{1, 2}}$;
in this way 
\begin{equation}
    \ke = \frac{\me}{M} \vb{P} - \vb{k}_1 - \vb{k}_2, \quad 
    \khi{1,2} = \vb{k}_{1, 2} + \frac{\mh}{M} \vb{P},
\end{equation}
and the kinetic energy part of the \eqref{eq:two-band-w} now reads 
\[
    H_{\text{kin}} = \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg 
    + \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mh} 
    - \frac{\vb{w}^2}{2M} + \frac{(\vb{k}_1 + \vb{k}_2 + \vb{w})^2}{2\me}.
\]
We however want to shift the definitions of $\vb{k}$ 
so that the kinetic part resembles the kinetic energy in a direct band gap system.
By redefining 
\begin{equation}
    \vb{k}_{1,2} = - \ii \hbar \partial_{\vb{r}_{1, 2}} + \frac{\vb{w}}{1 + \me / \mu}, \quad 
    \mu = \frac{\me \mh}{\me + \mh},
\end{equation}
and therefore 
\begin{equation}
    \begin{aligned}
        &\ke = \frac{\me}{M} \vb{P} + \frac{2 \mh \vb{w}}{M} - \vb{k}_1 - \vb{k}_2, \\ 
        &\khi{1,2} = \vb{k}_{1,2} - \frac{\mh \vb{w}}{M} + \frac{\mh}{M} \vb{P},
    \end{aligned}
\end{equation}
the total Hamiltonian now reads
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg, \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2}{2\mu} + \frac{\vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{2\me}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}}.
    \end{aligned}
    \label{eq:internal-ham}
\end{equation}
The redefinition of $\vb{k}_{1, 2}$ does not change the commutation relation 
between $\vb{r}_{1, 2}$ and $\vb{k}_{1, 2}$.
We find that the internal part of the Hamiltonian has the same form 
as the Hamiltonian in \cite{chang2021variationally}. 
This can also be seen as generalization of the indirect band gap exciton two-band model to the trion;
for the former we have \cite{rustagi2018photoemission} 
\begin{equation}
    H = \frac{(\vb{Q} - \vb{w})^2}{2M} + \Eg + \frac{\vb{k}^2}{2\mu} - V(\vb{r}), 
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        &M = \me + \me, \\
        &\ke = \frac{\me}{M} \vb{Q} + \vb{k} + \frac{\mh}{M} \vb{w}, \\
        &\kh = \frac{\mh}{M} \vb{Q} - \vb{k} - \frac{\mh}{M} \vb{w}.
    \end{aligned}
\end{equation}

The trion wave function therefore assumes the following form: 
\begin{equation}
    \psi_{S \vb{P}}(\vb{R}, \vb{r}_1, \vb{r}_2) = \ee^{\ii \vb{P} \cdot \vb{R}} \sum_{s_1, s_2} A_{s_1 s_2}^{S \vb{P}} \phi_{s_1}(\vb{r}_1) \phi_{s_2}(\vb{r}_2), 
\end{equation}
where $\phi_{s}(\vb{r})$ is an eigenfunction of the hydrogenic Hamiltonian $\frac{\vb{k}^2}{2\mu} - V(\vb{r})$, 
and $A_{s_1 s_2}^{S \vb{P}}$ reflects the mixing of the modes by interactive Hamiltonian $V(\vb{r}_1 - \vb{r}_2)$ and $\vb{k}_1 \cdot \vb{k}_2 / 2\me$.
Working in the momentum space we can also write 
\begin{equation}
    \psi_{S \vb{P}}(\vb{R}, \vb{p}_1, \vb{p}_2) = \ee^{\ii \vb{P} \cdot \vb{R}} \sum_{s_1, s_2} A_{s_1 s_2}^{S \vb{P}} \phi_{s_1}(\vb{p}_1) \phi_{s_2}(\vb{p}_2), 
\end{equation}
where $\phi_{s}(\vb{p})$ is the Fourier transform of $\phi_s(\vb{r})$.
The additional constraint required to pin down the relation between $\khi{1,2}$ and $\ke, \vb{P}$ is maximization of $\psi_{S \vb{P}}(\vb{R}, \vb{p}_1, \vb{p}_2)$.

@@

The shifts of the center and the curvature of ARPES signatures are even more clearly 
when the trion mode has two electrons and one hole.
A typical setting of such a trion mode can be described by the following two-band model, 
where the two electrons reside at different valleys:
\begin{equation}
    \begin{aligned}
    H &= \frac{\kei{1}^2}{2\me} + \frac{(\kei{2} - \vb{w})^2}{2\me} + \frac{\kh^2}{2\mh} \\
    &\quad + V(\rei{1} - \rei{2}) - V(\rh - \rei{1}) - V(\rh - \rei{2}).
    \end{aligned}
\end{equation}
The decomposition of the Hamiltonian into internal and external terms 
follows exactly the same procedure as in the two-peak model; 
we only need to swap the positions of electrons and holes. 
Therefore we get 
\begin{equation}
    \begin{aligned}
        \vb{r}_1 &= \rei{1} - \rh, \quad 
        \vb{r}_2 = \rei{2} - \rh, \\
        \vb{k}_1 &= - \ii \partial_{\vb{r}_1} + \frac{\me}{M} \vb{w}, \quad 
        \vb{k}_2 = - \ii \partial_{\vb{r}_2} - \frac{\me + \mh}{M} \vb{w}, 
    \end{aligned}
\end{equation}
and conversely 
\begin{equation}
    \begin{aligned}
        &\kh = \frac{\mh}{M} (\vb{P} - \vb{w}) - \vb{k}_1 - \vb{k}_2, \\
        &\kei{1} = \vb{k}_1 - \frac{\me}{M} \vb{w} + \frac{\me}{M} \vb{P}, \\
        &\kei{2} = \vb{k}_2 + \frac{\me + \mh}{M} \vb{w} + \frac{\me}{M} \vb{P} .
    \end{aligned}
\end{equation}
Now when one electron is driven out by the probe, 
it can be e1 or e2, so $\kp = \kei{1, 2}$, 
momentum conservation blocks any block terms between the two valleys, 
so we can evaluate the ARPES signature contributed by the two valleys separately.
This drastically changes the dispersion relation of the ARPES signature, 
but the structure of trion wave function is intact.
Therefore, the curvature of the ARPES signature can still be found by either postulating $\vb{k}_1 = \vb{k}_2$ or by postulating that one of the momenta vanishes.
When $\vb{k}_1 = \vb{k}_2$ and $\kp = \kei{1}$, we have 
\begin{equation}
    \kei{2} - \vb{w} = \kp, \quad \kh = \vb{P} - \vb{w} - 2 \kp, 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = E_{\text{bind}} - \frac{\kp^2}{2\me} - \frac{(2 \kp + \vb{w} - \vb{P})^2}{2\mh}, 
\end{equation}
while the effective mass is

>>>>

In most driven ARPES settings, the probe is much weaker than the pump 
and does not itself influence the electronic structure,
and the generic theory of ARPES can be formulated by 
treating the pump as a non-equilibrium self-energy, 
while the effect of the probe can be captured by the first order perturbation  
and hence the final ARPES intensity is proportional to the electron dipole and 
the pumped single-electron lesser Green function
\cite{freericks2009theoretical,schuler2021theory};
the theory reduced to the usual Fermi golden rule 
when the pump is turned off \cite{sobota2021angle,freericks2021two}.
When the pump pulse is strong at its peak but has faded away at the probing time, 
the main effect of pumping is preparing an initial state, 
and the ARPES intensity can be evaluated from 
pure-state first-order time dependent perturbation theory,
which gives a Fermi golden rule-like expression of the probability of an electron being pumped out of the sample~\cite{rustagi2018photoemission}.
Ref.~\cite{rustagi2018photoemission} is focused on excitons;
here we generalize its methodology to all excitations and have  
\begin{widetext}
    \begin{equation}
        I_{\vb{k}}(\omega, t) \propto \sum_{n_1, n_2} \sum_{n}
        \rho_n (M^{fn_2}_{\vb{k} \vb{k}_\parallel})^* M^{fn_1}_{\vb{k} \vb{k}_\parallel} 
        \int_{t_0}^t \dd{t_1} \int_{t_0}^t \dd{t_2}
        \ee^{-\ii (E_n - \omega ) (t_1 - t_2)}
        \mel{\Psi_n}{c^\dagger_{n_2 \vb{k}_\parallel} 
        U(t_2, t_1) c_{n_1 \vb{k}_\parallel}}{\Psi_n} s(t_1) s(t_2),
        \label{eq:exciton-response}
    \end{equation}
\end{widetext}
where $t$ is the observation time, 
$t_0$ is the starting time of the probe pulse,
$s(t)$ is the envelope function of the probe,
$\omega$ is the output electron energy shifted by the frequency of the probe pulse 
and the work function of the sample, 
$\vb{k}$ is the momentum of the outgoing electron 
and $\vb{k}_\parallel$ is its projection on the two-dimensional plane,  
$n_{1, 2}$ are electron band indices, 
$n = (S, \vb{Q})$ is the index of exciton modes, 
$\rho_n$ is the effective density matrix of excitons, 
$\ket*{\Psi_n}$ is the many-body wave function of the exciton state at $t=t_0$,
and $U(t_2, t_1)$ is the time evolution operator.
The $c^\dagger U c \ee^{- \ii E_n t}$ 
factor comes from the electron lesser Green function; 
the $\ee^{- \ii E_n t}$ factor comes from the time evolution of the excitation states, 
while the time evolution operator $U(t_2, t_1)$ 
instructs the time evolution of the residue of the excitation after one electron is driven out; 
time oscillation from the two factors together gives us the energy conservation relation
\begin{equation}
    \omega = E_n - E_{\text{residue}} , 
\end{equation}
where $E_{\text{residue}}$ refers to the remaining energy of the system after
one electron is driven out from $n_1, \kp$.
For three-dimensional systems, the momenta of $c$ and $c^\dagger$ in the $c^\dagger U c$ factor can be different \cite{freericks2009theoretical},  
but for two-dimensional materials however, since by momentum conservation  
the crystal momentum of the band electron that is driven out by the external field 
can only be the parallel component of $\vb{k}$, and we refer it as $\vb{k}_\parallel$ in \eqref{eq:exciton-response}.
As below we are mostly dealing with the ARPES spectrum of a single mode
In principle, the $\mel{\Psi_n}{c^\dagger_{n_2 \vb{k}_\parallel} U(t_2, t_1) c_{n_1 \vb{k}_\parallel}}{\Psi_n}$ matrix can be non-diagonal with an excited initial state;
frequently, however, because of energy conservation, the signature at given $(\omega, \vb{k})$
can only come from a single band, and in this case $n_1 = n_2$.

@@

We can define the canonical momenta of $\vb{r}_{1,2}$, 
hereafter referred to as $\vb{k}_{1,2}$, 
as $- \ii \hbar_{\vb{r}_{1, 2}}$;
in this way 
\begin{equation}
    \ke = \frac{\me}{M} \vb{P} - \vb{k}_1 - \vb{k}_2, \quad 
    \khi{1,2} = \vb{k}_{1, 2} + \frac{\mh}{M} \vb{P},
\end{equation}
and the kinetic energy part of the \eqref{eq:two-band-w} now reads 
\[
    H_{\text{kin}} = \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg 
    + \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mh} 
    - \frac{\vb{w}^2}{2M} + \frac{(\vb{k}_1 + \vb{k}_2 + \vb{w})^2}{2\me}.
\]
We however want to shift the definitions of $\vb{k}$ 
so that the kinetic part resembles the kinetic energy in a direct band gap system.
By redefining 
\begin{equation}
    \vb{k}_{1,2} = - \ii \hbar \partial_{\vb{r}_{1, 2}} + \frac{\vb{w}}{1 + \me / \mu}, \quad 
    \mu = \frac{\me \mh}{\me + \mh},
\end{equation}
and therefore 
\begin{equation}
    \begin{aligned}
        &\ke = \frac{\me}{M} \vb{P} + \frac{2 \mh \vb{w}}{M} - \vb{k}_1 - \vb{k}_2, \\ 
        &\khi{1,2} = \vb{k}_{1,2} - \frac{\mh \vb{w}}{M} + \frac{\mh}{M} \vb{P},
    \end{aligned}
\end{equation}
the total Hamiltonian now reads
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + \Eg \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2 + \vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{2\me}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}}.
    \end{aligned}
    \label{eq:internal-ham}
\end{equation}
The redefinition of $\vb{k}_{1, 2}$ does not change the commutation relation 
between $\vb{r}_{1, 2}$ and $\vb{k}_{1, 2}$.
We find that the internal part of the Hamiltonian has the same form 
as the trion Hamiltonian in \cite{chang2021variationally}, 
which includes no indirect band gap.
This can also be seen as generalization of the indirect band gap exciton two-band model to the trion;
for the exciton case we have \cite{rustagi2018photoemission} 
\begin{equation}
    H = \frac{(\vb{Q} - \vb{w})^2}{2M} + \Eg + \frac{\vb{k}^2}{2\mu} - V(\vb{r}), 
\end{equation}
where 
\begin{equation}
    \begin{aligned}
        &M = \me + \me, \\
        &\ke = \frac{\me}{M} \vb{Q} + \vb{k} + \frac{\mh}{M} \vb{w}, \\
        &\kh = \frac{\mh}{M} \vb{Q} - \vb{k} - \frac{\mh}{M} \vb{w}.
    \end{aligned}
\end{equation}

After diagonalization of \eqref{eq:internal-ham}, we get a trion spectrum with the form of 
\begin{equation}
    E_{S \vb{P}} = \frac{(\vb{P} - \vb{w})^2}{2M} + \Eg + E^{\text{b}}_{S},
\end{equation}
where $E_S^{\text{b}}$ comes from diagonalization of $H_{\text{internal}}$.
The trion wave function coming from diagonalization of  $H_{\text{internal}}$ assumes the following form: 
\begin{equation}
    \psi_{S \vb{P}}(\vb{R}, \vb{r}_1, \vb{r}_2) = \ee^{\ii \vb{P} \cdot \vb{R}} \sum_{s_1, s_2} A_{s_1 s_2}^{S \vb{P}} \phi_{s_1}(\vb{r}_1) \phi_{s_2}(\vb{r}_2), 
\end{equation}
where $\phi_{s}(\vb{r})$ is an eigenfunction of the hydrogenic Hamiltonian $\frac{\vb{k}^2}{2\mu} - V(\vb{r})$, 
and $A_{s_1 s_2}^{S \vb{P}}$ reflects the mixing of the modes by interactive Hamiltonian $V(\vb{r}_1 - \vb{r}_2)$ and $\vb{k}_1 \cdot \vb{k}_2 / 2\me$.
Working in the momentum space we can also write 
\begin{equation}
    \psi_{S \vb{P}}(\vb{R}, \vb{p}_1, \vb{p}_2) = \ee^{\ii \vb{P} \cdot \vb{R}} \sum_{s_1, s_2} A_{s_1 s_2}^{S \vb{P}} \phi_{s_1}(\vb{p}_1) \phi_{s_2}(\vb{p}_2), 
\end{equation}
where $\phi_{s}(\vb{p})$ is the Fourier transform of $\phi_s(\vb{r})$.
The additional constraint required to pin down the relation between $\khi{1,2}$ and $\ke, \vb{P}$ is maximization of $\psi_{S \vb{P}}(\vb{R}, \vb{p}_1, \vb{p}_2)$.

@@

The shifts of the center and the curvature of ARPES signatures are even more clearly 
when the trion mode has two electrons and one hole.
A typical setting of such a trion mode can be described by the following two-band model, 
where the two electrons reside at different valleys:
\begin{equation}
    \begin{aligned}
    H &= \frac{\kei{1}^2}{2\me} + \frac{(\kei{2} - \vb{w})^2}{2\me} + \frac{\kh^2}{2\mh} + 2 \Eg \\
    &\quad + V(\rei{1} - \rei{2}) - V(\rh - \rei{1}) - V(\rh - \rei{2}).
    \end{aligned}
\end{equation}
The decomposition of the Hamiltonian into internal and external terms 
follows exactly the same procedure as in the two-peak model; 
we only need to swap the positions of electrons and holes. 
Therefore we find that under 
\begin{equation}
    \begin{aligned}
        \vb{r}_1 &= \rei{1} - \rh, \quad 
        \vb{r}_2 = \rei{2} - \rh, \\
        \vb{k}_1 &= - \ii \partial_{\vb{r}_1} + \frac{\me}{M} \vb{w}, \quad 
        \vb{k}_2 = - \ii \partial_{\vb{r}_2} - \frac{\me + \mh}{M} \vb{w}, 
    \end{aligned}
\end{equation}
or conversely 
\begin{equation}
    \begin{aligned}
        &\kh = \frac{\mh}{M} (\vb{P} - \vb{w}) - \vb{k}_1 - \vb{k}_2, \\
        &\kei{1} = \vb{k}_1 - \frac{\me}{M} \vb{w} + \frac{\me}{M} \vb{P}, \\
        &\kei{2} = \vb{k}_2 + \frac{\me + \mh}{M} \vb{w} + \frac{\me}{M} \vb{P} ,
    \end{aligned}
\end{equation}
the Hamiltonian of the two-electron-one-hole trion is reduced to 
% See src/two-electron.nb
\begin{equation}
    \begin{aligned}
        H &=  \frac{(\vb{P} - \vb{w})^2}{2 M} + 2 \Eg, \\ 
        &\ \  + \underbrace{
            \frac{\vb{k}_1^2}{2\mu} + \frac{\vb{k}_2^2}{2\mu} + \frac{\vb{k}_1 \cdot \vb{k}_2}{2\mh}
            + V(\vb{r}_1 - \vb{r}_2) - V(\vb{r}_1) - V(\vb{r}_2)
        }_{H_{\text{internal}}},
    \end{aligned}
\end{equation}
a form resembling \eqref{eq:internal-ham}, 
the only differences being the coefficient of the $\vb{k}_1 \cdot \vb{k}_2$ term
and the $2\Eg$ term due to the fact that there are two electrons now.
Now when one electron is driven out by the probe, 
it can be e1 or e2, so $\kp = \kei{1,2}$, 
momentum conservation blocks any block terms between the two valleys, 
so we can evaluate the ARPES signature contributed by the two valleys separately.
This drastically changes the dispersion relation of the ARPES signature, 
but the structure of trion wave function is intact.
The most remarkable difference between the ARPES signature of a negative trion 
and that of a positive trion
is the former has a vertical ``tail'': 
when $\kp = \kei{1}$ is chosen, $\vb{k}_2$ is allow to move freely, 
and it continuously adjusts the energy of the remaining of the trion after e1 is driven out, 
which means the ARPES signature will have a long, vertical component 
that does not coincide with any dispersion relations. % TODO: rework the wording

Therefore, the curvature of the ARPES signature can still be found by either postulating $\vb{k}_1 = \vb{k}_2$ or by postulating that one of the momenta vanishes.
When $\vb{k}_1 = \vb{k}_2$ and $\kp = \kei{1}$, we have 
\begin{equation}
    \kei{2} - \vb{w} = \kp, \quad \kh = \vb{P} - \vb{w} - 2 \kp, 
\end{equation}
and the dispersion relation is 
\begin{equation}
    \omega = \Eg + E_{\text{b}} + \frac{\vb{P}^2}{2M} - \frac{\kp^2}{2\me} - \frac{(2 \kp + \vb{w} - \vb{P})^2}{2\mh}, 
\end{equation}
while the effective mass is
